---
title: "ML Dumbell"
author: "Scott Dayer"
date: "5/18/2019"
output: html_document
---

## Executive Summary

This report describes the process and analysis performed to build a model that accurately predicts the method ("classe") in which a dumbell is raised from the Weight Lifting Exercise Dataset. After exploratory data analysis is conducted, and pre-processing strategies are identified, several models are constructed and analyzed on a validation set of data, and a final model is selected for optimal accuracy and processing speed for use on the test data in order to submit the course quiz.  

The selected model utilizes principal component analysis for pre-processing, k-folds cross validation, a random forest machine learning algorithm.  Identification of the correct "classe" is approximately 99% accurate.  

```{r}

library(tidyverse)
library(caret)
library(doParallel)

training <- as_tibble(read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")))
testing <- as_tibble(read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")))

```
## Exploratory Data Analysis


```{r}

head()

na_count <- sapply(training, function(y) sum(length(which(is.na(y)))))
na_count <- data.frame(na_count)

#identify the variables with near zero variance
# to eliminate certain variables that 
nzv <- nearZeroVar(training)

#remove all variable with near zero variance
trainingNNZV <- training[,-c(nzv)]
100
#remove all columns with NAs
training2 <- trainingNNZV[,colSums(is.na(trainingNNZV)) < 15000]
t <- training2[,6:59]

# create a validation set from training set

inTrain <- createDataPartition(t$classe, p = .9, list = FALSE)
trainSet <- t[inTrain,]
validationSet <- t[-inTrain,]

# correlated predictors


M <- (abs(cor(t)))
diag(M) <- 0
which(M > 0.8, arr.ind = T)

# PCA

preProc <- preProcess(training2, method = "pca")
trainPC <- predict(preProc, training2)
modelFit <- train(training2$classe ~ ., method = "rf", data = trainPC)

# exploratory charts


        
        
#featurePlot(x = training2, y = training2$classe, plot = "pairs")
```

# k-fold cross validation

```{r}

kf3 <- trainControl(method = "cv", number = 3, allowParallel = TRUE)
kf7 <- trainControl(method = "cv", number = 7, allowParallel = TRUE)


```

#Validation Testing
```{r}


cluster <- makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)

model_gbm <- train(classe ~ ., method = "gbm", data = trainSet, preProcess = "pca")
model_rf <- train(classe ~ ., method = "rf", data = trainSet, preProcess = "pca")
model_rfcv <- train(classe ~ ., method = "rf", data = trainSet, preProcess = "pca",
                    trControl = kf3)
model_rfcv7 <- train(classe ~ ., method = "rf", data = trainSet, preProcess = "pca",
                    trControl = kf7)

print(model_gbm)
pred_gbm <- predict(model_gbm, validationSet)
gbm_resultDF <- data.frame(pred_gbm, actual = validationSet$classe) 
mean(pred_gbm == validationSet$classe)

print(model_rf)
pred_rf <- predict(model_rf, validationSet)
rf_resultDF <- data.frame(pred_rf, actual = validationSet$classe)
mean(pred_rf == validationSet$classe)

print(model_rfcv)
pred_rfcv <- predict(model_rfcv, validationSet)
rfcv_resultDF <- data.frame(pred_rfcv, actual = validationSet$classe)
mean(pred_rfcv == validationSet$classe)

print(model_rfcv7)
pred_rfcv7 <- predict(model_rfcv7, validationSet)
rfcv7_resultDF <- data.frame(pred_rfcv7, actual = validationSet$classe)
mean(pred_rfcv7 == validationSet$classe)

stopCluster(cluster)
registerDoSEQ()

```

# Prediction quiz

```{r}

pred_gbm_quiz <- predict(model_gbm, testing)
pred_rf_quiz <- predict(model_rf, testing)
```


## 5 different categories:

Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: 

exactly according to the specification (Class A) 
throwing the elbows to the front (Class B) 
lifting the dumbbell only halfway (Class C) 
lowering the dumbbell only halfway (Class D)
throwing the hips to the front (Class E)

